Formalizing the Edmonds-Karp Algorithm

### 2016-03-02 paper/root.tex

\section{Introduction}
The Ford-Fulkerson algorithm [cite] is one of the important results of the graph theory and is used to find a solution to the maximum flow problem in flow networks. Many important mathematical problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and many other scheduling and resource allocating problems can be reduced to the maximum flow problem. Hence, the Ford-Fulkerson algorithm has major application in the field of mathematical optimization.

Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants. The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar [cite]. This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm. Providing such an implementation is specially important, as it could provide the basis for many verfied programs with practical importance.

This paper present a formalization of the correctness proof of the Ford-Fulkerson algorithm in Isabelle/HOL. Our proof is based on the informal proof of the algorithm which is presented in the book "Introduction to algorithms" [cite]. Due to practical importance of the algorithm, we also present a verified implementation of the algorithm. Isabelle/HOL provides some automation for generating the code corresponding to a verified algorithm, however, in order to use the code generating features, one needs to do the formalization with executability in mind. Being limited only to executable concepts in the formalization has the disadvantage of cluttering the proofs with implementation details. Such approach makes the proofs more complicated, and may even render proofs of medium complex algorithms unmanageable.

One solution for the aforementioned problem is refinement [cite]. In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm. Next, we refine the abstract definition of the algorithm towards an executable implementation in possibly multiple refinement steps. During each step, we only need to prove the correctness of the implementation of a particular abstract concept. Hence, we can be sure about the correctness of the resulting executable program as each refinement step preserves the correctness.

There are several approaches to data refinement in Isabelle/HOL. We will be using the Autoref tool [cite], which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm [cite]. Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem. To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework [cite] and the Isabelle Collection Framework [cite].

@@

Informal proofs focus on the relevant thoughts by leaving out technical details. However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow. In the case that an interactive theorem prover is used, readers may even need to replay the proof to understand the idea behind it. In Isabelle proof assistant, the Isar proof language bridges the semantic gap between internal notions of proof given by interactive theorem proving systems and an appropriate level of abstraction for user-level work. Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts. We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al. [cite].

@@

In reference books, it is a convention to use pseudocodes for presenting the algorithms. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm and leaves out the implementation details. We present the Ford-Fulkerson method using the syntax of Autoref[cite] tool which is quite similar to pseudocode of the algorithm:
\Snippet{ford_fulkerson_algo}

>>>>

\section{Introduction}
The Ford-Fulkerson algorithm~\cite{FF56} is one of the important results of the graph theory and is used to find a solution to the maximum flow problem in flow networks. Many important mathematical problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and many other scheduling and resource allocating problems can be reduced to the maximum flow problem. Hence, the Ford-Fulkerson algorithm has major application in the field of mathematical optimization.

Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants. The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}. This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm. Providing such an implementation is specially important, as it could provide the basis for many verfied programs with practical importance.

This paper present a formalization of the correctness proof of the Ford-Fulkerson algorithm in Isabelle/HOL~\cite{NPW02}. Our proof is based on the informal proof of the algorithm which is presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}. Due to practical importance of the algorithm, we also present a verified implementation of the algorithm. Isabelle/HOL provides some automation for generating the code corresponding to a verified algorithm, however, in order to use the code generating features, one needs to do the formalization with executability in mind. Being limited only to executable concepts in the formalization has the disadvantage of cluttering the proofs with implementation details. Such approach makes the proofs more complicated, and may even render proofs of medium complex algorithms unmanageable.

One solution for the aforementioned problem is refinement~\cite{Morr87}. In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm. Next, we refine the abstract definition of the algorithm towards an executable implementation in possibly multiple refinement steps. During each step, we only need to prove the correctness of the implementation of a particular abstract concept. Hence, we can be sure about the correctness of the resulting executable program as each refinement step preserves the correctness.

There are several approaches to data refinement in Isabelle/HOL. We will be using the Autoref tool~\cite{La13}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}. Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem. To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework~\cite{La12} and the Isabelle Collection Framework~\cite{LL10}.

@@

Informal proofs focus on the relevant thoughts by leaving out technical details. However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow. In the case that an interactive theorem prover is used, readers may even need to replay the proof to understand the idea behind it. In Isabelle proof assistant, the Isar proof language bridges the semantic gap between internal notions of proof given by interactive theorem proving systems and an appropriate level of abstraction for user-level work. Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts. We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.~\cite{CLRS09}.

@@

In reference books, it is a convention to use pseudocodes for presenting the algorithms. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm and leaves out the implementation details. We present the Ford-Fulkerson method using the syntax of Autoref~\cite{La13} tool which is quite similar to pseudocode of the algorithm:
\Snippet{ford_fulkerson_algo}
